---
title: "Gata2_affy_pre_process"
author: "Melanie Smith"
date: "2026-04-21"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

# Clean-up environment

```{r}
# clear all objects including hidden objects
rm(list = ls(all.names = TRUE))
# free up ram
gc()
# set some default options
options(
  max.print        = .Machine$integer.max,
  scipen           = 999,
  stringsAsFactors = FALSE,
  dplyr.summarise.inform = FALSE
)
# set seed for reproducibility
set.seed(42)
```

# Set project directories and file paths

```{r}
project_name     <- "20260421_natashaHarvey_gata2_mouse"
base_repo_dir    <- "/home/melanie-smith/workDir/natashaHarvey"

project_repo_dir   <- file.path(base_repo_dir, project_name)
raw_data_dir       <- file.path(project_repo_dir, "data", "raw")
processed_data_dir <- file.path(project_repo_dir, "data", "processed")  # consistent name throughout
out_dir            <- file.path(project_repo_dir, "outDir")
qc_dir             <- file.path(out_dir, "QC")

# create the output directories if they don't already exist
for (d in c(processed_data_dir, out_dir, qc_dir)) {
  dir.create(d, recursive = TRUE, showWarnings = FALSE)
  cat("Output directory confirmed:", d, "\n")
}
```

# Load libraries

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

required_pkgs <- c("oligo",
                   "pd.mogene.1.0.st.v1",
                   "oligoClasses",  # affyPLM-equivalent probe-level modelling via oligo
                   "limma",
                   "RColorBrewer",
                   "ggplot2",
                   "ggrepel",
                   "pheatmap")

for (pkg in required_pkgs) {
  # install packages if not already installed
  if (!requireNamespace(pkg, quietly = TRUE)) {
    BiocManager::install(pkg, ask = FALSE, update = FALSE)
  }
  # load the required libraries
  library(pkg, character.only = TRUE)
}
```

# Locate and validate CEL files

```{r}
# list files in the raw data directory
cel_files <- list.celfiles(raw_data_dir, full.names = TRUE)
# ensure reproducible ordering
cel_files <- sort(cel_files)
# stop if there are no files found
if (length(cel_files) == 0) {
  stop("No CEL files found in: ", raw_data_dir)
}

cat("Found", length(cel_files), "CEL files:\n")
cat(paste0("  ", basename(cel_files), collapse = "\n"), "\n")

# Validate array type from CEL headers before reading to avoid cryptic errors
# from stray files of a different platform landing in the raw directory.
# affxparser::readCelHeader is the correct function — oligo depends on it
# internally but does not export readCelHeader in recent Bioconductor releases.
if (!requireNamespace("affxparser", quietly = TRUE))
  BiocManager::install("affxparser", ask = FALSE, update = FALSE)

cel_headers <- lapply(cel_files, affxparser::readCelHeader)
array_types  <- sapply(cel_headers, `[[`, "chiptype")
cat("\nArray types detected:\n")
print(table(array_types))

if (length(unique(array_types)) > 1) {
  stop("Mixed array types detected — all CEL files must be from the same platform.\n",
       "Types found: ", paste(unique(array_types), collapse = ", "))
}
```

# Read CEL files

```{r}
raw_data <- oligo::read.celfiles(cel_files)

# Strip path and .CEL extension, then extract "group_replicate" (fields 2 and 3)
sampleNames(raw_data) <- sapply(
  tools::file_path_sans_ext(basename(cel_files)),
  function(x) paste(strsplit(x, "_")[[1]][2:3], collapse = "_")
)

cat("\nSample names after cleaning:\n")
print(sampleNames(raw_data))
```

# Build sample metadata

```{r}
# pData() from an oligo read.celfiles() object contains only file paths by default — not useful as a metadata file on its own.
# Here I will build a structured metadata table and add group/condition columns to match the experimental design before proceeding.

sample_metadata <- data.frame(
  sample_name = sampleNames(raw_data),
  cel_file    = basename(cel_files),
  group       = c("con","con","con","con","GATA2","GATA2","GATA2","GATA2"),
  stringsAsFactors = FALSE,
  row.names   = sampleNames(raw_data)   # <-- this is what was missing
)

pData(raw_data) <- sample_metadata

cat("\nSample metadata:\n")
print(pData(raw_data))
```

# ── PRE-NORMALISATION QC ──────────────────────────────────────────────────────

## Raw intensity boxplots

```{r fig.cap="Boxplots of raw log2 probe intensities. Boxes should be roughly similar in spread and position. Outlier arrays will appear visually shifted or with markedly different spread." fig.keep="all}
# log2 raw intensities (oligo stores on log2 scale already via exprs on raw)
raw_exprs <- log2(pm(raw_data))

# Colour by sample for visual separation
n_samples <- ncol(raw_exprs)
pal       <- colorRampPalette(RColorBrewer::brewer.pal(8, "Set2"))(n_samples)

png(file.path(out_dir, "QC/raw_probe_intensities_boxplot.png"),
    width  = 2400,
    height = 1200,
    res    = 300)
par(mar = c(9, 4, 3, 1))
boxplot(raw_exprs,
        col     = pal,
        las     = 2,
        cex.axis = 0.7,
        main    = "Raw probe intensities (log2 PM)",
        ylab    = "log2 intensity",
        outline = FALSE)
abline(h = median(raw_exprs, na.rm = TRUE), col = "red", lty = 2, lwd = 1.5)
dev.off()
```

## Raw intensity density plots

```{r fig.cap="Density plots of raw log2 intensities per array. All curves should overlap closely. Shifted or multi-modal curves indicate a problem with that sample." fig.keep="all"}

png(file.path(out_dir, "QC/raw_probe_intensities_density.png"),
    width  = 1800,
    height = 1200,
    res    = 300)
par(mar = c(5, 4, 3, 8), xpd = TRUE)
plot(density(raw_exprs[, 1]),
     main = "Raw intensity density (log2 PM)",
     xlab = "log2 intensity",
     ylim = c(0, 0.25),
     col  = pal[1],
     lwd  = 1.5)
for (i in seq_len(n_samples)[-1]) {
  lines(density(raw_exprs[, i]), col = pal[i], lwd = 1.5)
}
legend("topright",
       inset   = c(-0.25, 0),
       legend  = colnames(raw_exprs),
       col     = pal,
       lwd     = 1.5,
       cex     = 0.65,
       bty     = "n")
dev.off()
```

## RNA degradation plot

```{r fig.cap="RNA degradation plot. Each line represents one sample. Lines should be broadly parallel and flat. Steep positive slopes (5' to 3') indicate RNA degradation in that sample."}
# oligo provides a degradation-style plot via image() and boxplot on raw
# For MoGene ST arrays the full AffyRNAdeg() isn't available, but we can
# approximate with the built-in oligo image for spatial artefacts, and use
# the 3'/5' probe-index intensity trend manually.
#
# The oligo::image() call below plots a spatial heatmap for each array.
# Run it in a loop and check for spatial gradients (blobs, stripes, edges).
cat("Generating spatial artefact images for each array...\n")
for (i in seq_len(ncol(raw_data))) {
  png(file.path(qc_dir, paste0("spatial_image_", sampleNames(raw_data)[i], ".png")),
      width  = 1200,
      height = 1200,
      res    = 300)
  
  oligo::image(raw_data[, i],
               main = paste("Spatial image:", sampleNames(raw_data)[i]))
  
  dev.off()
}
```

## PCA — post-normalisation

```{r fig.cap="PCA of raw log2 expression. Samples will cluster by whatever factor contains the most variance."}
# note raw_exprs is log2 data
pca_res_raw <- prcomp(t(raw_exprs), scale. = FALSE)
pca_df_raw  <- as.data.frame(pca_res_raw$x[, 1:2])
pca_df_raw$sample <- rownames(pca_df_raw)
pca_df_raw$group  <- sample_metadata$group[match(pca_df_raw$sample,
                                             sample_metadata$sample_name)]

# Variance explained
pct_var_raw <- round(100 * pca_res_raw$sdev^2 / sum(pca_res_raw$sdev^2), 1)

pca_plot_raw <- ggplot(pca_df_raw,
                   aes(x = PC1,
                       y = PC2,
                       colour = group,
                       label = sample)) +
  geom_point(size = 4, alpha = 0.9) +
  ggrepel::geom_text_repel(size = 3.2,
                           show.legend = FALSE,
                           max.overlaps = Inf) +  # requires ggrepel
  scale_colour_manual(values = c("con" = "#003087", "GATA2" = "#CC0000")) +
  labs(
    title  = "PCA — raw log2 expression",
    x      = paste0("PC1 (", pct_var_raw[1], "% variance)"),
    y      = paste0("PC2 (", pct_var_raw[2], "% variance)"),
    colour = "Group"
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "right")
ggsave(plot = pca_plot_raw,
       filename = file.path(out_dir, "pca_raw_1_2.png"),
                            width = 8,
                            height = 6,
                            units = "in")
```

# RMA normalisation without prior normalisation

```{r}
# perform RMA without prior normalisation
rma_no_norm_data <- oligo::rma(raw_data, target = "core", normalize = FALSE)

cat("\nRMA normalisation complete.\n")
cat("Dimensions (probes × samples):", dim(exprs(rma_no_norm_data)), "\n")
```

## Relative Log Expression (RLE) plot — pre-normalisation

```{r fig.cap="RLE plot of raw data. Each box shows the distribution of log2 expression values for one array after subtracting the row-wise median across arrays. Boxes should be centred near zero and compact. Arrays with shifted medians or wide spreads are candidates for investigation or exclusion." fig.keep="all"}
# Compute RLE: subtract row median across all arrays from each value
row_medians <- Biobase::rowMedians(as.matrix(Biobase::exprs(rma_no_norm_data)))
rle_mat     <- sweep(as.matrix(Biobase::exprs(rma_no_norm_data)), 1, row_medians, "-")

png(file.path(out_dir, "QC/raw_probe_RLE_boxplot.png"),
    width  = 2400,
    height = 1200,
    res    = 300)
par(mar = c(9, 4, 3, 1))
oligo::boxplot(rle_mat,
        col      = pal,
        las      = 2,
        cex.axis = 0.7,
        main     = "RLE plot — raw data (RMA no normalisation)",
        ylab     = "Relative log2 expression", # log2_expression_deviation
        outline  = FALSE)
abline(h = 0, col = "red", lty = 2, lwd = 1.5)
dev.off()

# Summarise RLE medians as a QC table
rle_summary <- data.frame(
  sample     = colnames(rle_mat),
  rle_median = round(apply(rle_mat, 2, median, na.rm = TRUE), 4),
  rle_iqr    = round(apply(rle_mat, 2, IQR,    na.rm = TRUE), 4)
)
cat("\nRLE summary (medians should be ≈ 0; IQR should be small and consistent):\n")
print(rle_summary)

# Flag potential outliers: RLE median > 0.1 or IQR > 1.5× the group median IQR
median_iqr  <- median(rle_summary$rle_iqr)
flagged     <- rle_summary[abs(rle_summary$rle_median) > 0.1 |
                            rle_summary$rle_iqr > 1.5 * median_iqr, ]
if (nrow(flagged) > 0) {
  warning("Potential outlier arrays flagged by RLE:\n",
          paste(flagged$sample, collapse = ", "),
          "\nReview these samples carefully before proceeding.")
} else {
  cat("No outlier arrays flagged by RLE criteria.\n")
}
```

# RMA normalisation with prior normalisation

```{r}
# perform RMA with prior normalisation
rma_norm_data <- oligo::rma(raw_data, target = "core", normalize = TRUE)

cat("\nRMA normalisation complete.\n")
cat("Dimensions (probes × samples):", dim(exprs(rma_norm_data)), "\n")
```

# Mean SD plot

```{r}

# Mean-SD plot
plotMeanSD <- function(mat) {
  row_means <- rowMeans(mat)
  row_sds   <- apply(mat, 1, sd)
  
  plot(row_means, row_sds,
       pch  = 16,
       cex  = 0.3,
       col  = adjustcolor("steelblue", alpha.f = 0.4),
       xlab = "Mean log2 expression",
       ylab = "Standard deviation",
       main = "Mean-SD plot — RMA normalised")
  
  # Add a loess smoothed line to show the trend
  loess_fit <- loess(row_sds ~ row_means)
  lines(sort(row_means),
        predict(loess_fit)[order(row_means)],
        col = "red", lwd = 2)
}

plotMeanSD(exprs(rma_norm_data))

```

# Extract expression matrix

```{r}
expr_matrix <- exprs(rma_norm_data)

# FIX: add a dimension check here to catch silent problems early
if (nrow(expr_matrix) == 0 | ncol(expr_matrix) == 0) {
  stop("Expression matrix is empty after RMA — investigate read.celfiles() output.")
}
cat("Expression matrix dimensions:", nrow(expr_matrix), "probes ×",
    ncol(expr_matrix), "samples\n")
cat("Expression value range: [",
    round(min(expr_matrix), 2), ",",
    round(max(expr_matrix), 2), "]\n")
```

# ── POST-NORMALISATION QC ─────────────────────────────────────────────────────

## Normalised intensity boxplots

```{r fig.cap="Boxplots of RMA-normalised log2 expression values. After normalisation all boxes should be tightly aligned at the same median with similar spread. Persistent outliers after RMA warrant investigation." fig.keep="all"}

png(file.path(out_dir, "QC/normalised_probe_RLE_boxplot.png"),
    width  = 2400,
    height = 1200,
    res    = 300)
par(mar = c(9, 4, 3, 1))
oligo::boxplot(expr_matrix,
               col      = pal,
               las      = 2,
               cex.axis = 0.7,
               main     = "RMA-normalised expression (log2)",
               ylab     = "log2 expression",
               outline  = FALSE)
abline(h = median(expr_matrix, na.rm = TRUE), col = "red", lty = 2, lwd = 1.5)
dev.off()
```

## Normalised density plots

```{r fig.cap="Density plots of RMA-normalised expression. All distributions should now overlap almost perfectly. Persistent separation after RMA is unusual and should be investigated."}

png(file.path(out_dir, "QC/normalised_probe_RLE_density.png"),
    width  = 1800,
    height = 1200,
    res    = 300)
par(mar = c(5, 4, 3, 8), xpd = TRUE)
plot(density(expr_matrix[, 1]),
     main = "Normalised expression density (log2)",
     xlab = "log2 expression",
     col  = pal[1],
     lwd  = 1.5)
for (i in seq_len(n_samples)[-1]) {
  lines(density(expr_matrix[, i]), col = pal[i], lwd = 1.5)
}
legend("topright",
       inset  = c(-0.25, 0),
       legend = colnames(expr_matrix),
       col    = pal,
       lwd    = 1.5,
       cex    = 0.65,
       bty    = "n")
dev.off()
```

## PCA — post-normalisation

```{r fig.cap="PCA of RMA-normalised expression. Samples should cluster by biological group. Clustering by batch, processing date, or technician indicates a confound that may need to be addressed (e.g. ComBat correction). Label each point with its sample name to identify any outlier positions."}
pca_res <- prcomp(t(expr_matrix), scale. = FALSE)
pca_df  <- as.data.frame(pca_res$x[, 1:2])
pca_df$sample <- rownames(pca_df)
pca_df$group  <- sample_metadata$group[match(pca_df$sample,
                                             sample_metadata$sample_name)]

# Variance explained
pct_var <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2), 1)

pca_plot <- ggplot(pca_df,
                   aes(x = PC1,
                       y = PC2,
                       colour = group,
                       label = sample)) +
  geom_point(size = 4, alpha = 0.9) +
  ggrepel::geom_text_repel(size = 3.2,
                           show.legend = FALSE,
                           max.overlaps = Inf) +  # requires ggrepel
  scale_colour_manual(values = c("con" = "#003087", "GATA2" = "#CC0000")) +
  labs(
    title  = "PCA — RMA-normalised expression",
    x      = paste0("PC1 (", pct_var[1], "% variance)"),
    y      = paste0("PC2 (", pct_var[2], "% variance)"),
    colour = "Group"
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "right")
ggsave(plot = pca_plot,
       filename = file.path(out_dir, "pca_norm_1_2.png"),
                            width = 8,
                            height = 6,
                            units = "in")
```

## Sample-to-sample correlation heatmap
## Pearson's correlation

```{r fig.cap="Pairwise Pearson correlation heatmap of RMA-normalised expression. All values should be high (> 0.95 for good microarray data). Samples with systematically lower correlations to all others are candidates for exclusion."}
cor_mat <- cor(expr_matrix, method = "pearson")

# Annotate rows/cols with group
ann_df <- data.frame(Group = sample_metadata$group,
                     row.names = sample_metadata$sample_name)

pheatmap(cor_mat,
         annotation_col  = ann_df,
         annotation_row  = ann_df,
         color           = colorRampPalette(c("#2166AC", "white", "#B2182B"))(100),
         breaks          = seq(0.85, 1, length.out = 101),
         display_numbers = TRUE,
         number_format   = "%.3f",
         fontsize_number = 7,
         main            = "Sample-to-sample Pearson correlation",
         border_color    = NA,
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean")

# Flag any pairwise correlation below threshold
cor_threshold <- 0.95
low_cor       <- which(cor_mat < cor_threshold & upper.tri(cor_mat), arr.ind = TRUE)
if (nrow(low_cor) > 0) {
  cat("\nSample pairs with Pearson r <", cor_threshold, ":\n")
  for (k in seq_len(nrow(low_cor))) {
    cat(sprintf("  %s vs %s: r = %.4f\n",
                rownames(cor_mat)[low_cor[k, 1]],
                colnames(cor_mat)[low_cor[k, 2]],
                cor_mat[low_cor[k, 1], low_cor[k, 2]]))
  }
} else {
  cat("All pairwise Pearson correlations are >=", cor_threshold, "\n")
}
```

## Manhattan Distance

```{r}
dists <- as.matrix(dist(t(expr_matrix), method = "manhattan"))
rownames(dists) <- row.names(pData(rma_norm_data))
hmcol <- rev(colorRampPalette(RColorBrewer::brewer.pal(9, "YlOrRd"))(255))

# Annotate rows/cols with group
ann_df <- data.frame(Group = sample_metadata$group,
                     row.names = sample_metadata$sample_name)

pheatmap(dists, col = (hmcol),
         annotation_col  = ann_df,
         annotation_row  = ann_df,
         legend = TRUE,
         treeheight_row = 0,
         display_numbers = TRUE,
         number_format   = "%.3f",
         legend_breaks = c(min(dists, na.rm = TRUE),
                           max(dists, na.rm = TRUE)),
         legend_labels = (c("small distance", "large distance")),
         main = "Clustering heatmap for the calibrated samples")
```

## RLE plot — post-normalisation

```{r fig.cap="RLE plot post-RMA. After normalisation boxes should be tightly centred on zero with minimal spread. Any remaining wide spread or shifted median warrants investigation."}
row_medians_norm <- apply(expr_matrix, 1, median, na.rm = TRUE)
rle_mat_norm     <- sweep(expr_matrix, 1, row_medians_norm, "-")

png(file.path(out_dir, "QC/normalised_probe_RLE_boxplot.png"),
    width  = 2400,
    height = 1200,
    res    = 300)
par(mar = c(9, 4, 3, 1))
oligo::boxplot(rle_mat_norm,
        col      = pal,
        las      = 2,
        cex.axis = 0.7,
        main     = "RLE plot — post-RMA normalisation",
        ylab     = "Relative log2 expression",
        outline  = FALSE)
abline(h = 0, col = "red", lty = 2, lwd = 1.5)
dev.off()

```

# Intensity level filtering 

```{r}
# ── Step 1: Compute per-probe medians and plot histogram ──────────────────────
probe_medians <- rowMedians(Biobase::exprs(rma_norm_data))

hist(probe_medians,
     100,
     col    = "cornsilk1",
     freq   = FALSE,
     main   = "Histogram of median probe intensities (RMA normalised)",
     border = "antiquewhite4",
     xlab   = "Median log2 intensity")

# ── Step 2: Set threshold visually ───────────────────────────────────────────
# Inspect the histogram output above, then set man_threshold to the left
# of the main peak. Adjust this value after inspecting your histogram.
man_threshold <- 5.0   # <-- update after inspecting histogram

# Overlay threshold line on histogram
hist(probe_medians,
     100,
     col    = "cornsilk1",
     freq   = FALSE,
     main   = "Histogram of median probe intensities (RMA normalised)",
     border = "antiquewhite4",
     xlab   = "Median log2 intensity")
abline(v = man_threshold, col = "coral4", lwd = 2)

# ── Step 3: Determine samples cutoff from smallest group ─────────────────────
# For your dataset: n=4 con, n=4 GATA2 — smallest group is 4
no_of_samples  <- table(pData(norm_data)$group)
samples_cutoff <- min(no_of_samples)

cat("Group sizes:\n")
print(no_of_samples)
cat("Samples cutoff (smallest group size):", samples_cutoff, "\n")

# ── Step 4: Apply per-sample filter ──────────────────────────────────────────
# Keep probe if intensity > threshold in at least `samples_cutoff` samples
idx_man_threshold <- apply(
  Biobase::exprs(norm_data), 1,
  function(x) sum(x > man_threshold) >= samples_cutoff
)

cat("\nProbe filter summary:\n")
print(table(idx_man_threshold))

# ── Step 5: Subset the normalised ExpressionSet ───────────────────────────────
norm_filtered <- subset(norm_data, idx_man_threshold)

cat("\nProbes before filtering:", nrow(norm_data),    "\n")
cat("Probes after filtering: ", nrow(norm_filtered), "\n")
cat("Probes removed:         ", nrow(norm_data) - nrow(norm_filtered), "\n")

```

# Save outputs

```{r}
write.csv(expr_matrix,
          file  = file.path(processed_data_dir, "expression_matrix_rma.csv"),
          quote = FALSE)

saveRDS(expr_matrix,
        file = file.path(processed_data_dir, "expression_matrix_rma.rds"))

saveRDS(norm_data,
        file = file.path(processed_data_dir, "norm_data_rma.rds"))

# Save filtered ExpressionSet for downstream use
saveRDS(norm_filtered, file.path(processed_data_dir, "norm_data_filtered.rds"))

cat("Expression matrix saved:", nrow(expr_matrix), "probes ×",
    ncol(expr_matrix), "samples\n")
```

# Save sample metadata

```{r}
# FIX: was referencing 'processed_data' (undefined); corrected to processed_data_dir
# FIX: now saves the structured metadata table rather than the sparse pData() output
write.csv(sample_metadata,
          file  = file.path(processed_data_dir, "sample_metadata.csv"),
          quote = FALSE,
          row.names = FALSE)

cat("Sample metadata saved.\n")
```

# Session info for reproducibility

```{r}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo_preprocessing.txt"))
cat("Processing complete.\n")
```
