---
title: "probe_to_gene_mapping"
author: "Melanie Smith"
date: "2026-04-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Clean-up environment

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(
  max.print = .Machine$integer.max,
  scipen = 999,
  stringsAsFactors = FALSE,
  dplyr.summarise.inform = FALSE
)
set.seed(42) # set seed for reproducibility
```

# Set project directories and file paths

```{r}
project_name  <- "20260421_natashaHarvey_gata2_mouse"
base_repo_dir <- "/home/melanie-smith/workDir/natashaHarvey"

project_repo_dir <- file.path(base_repo_dir, project_name)
# raw_data_dir         <- file.path(project_repo_dir, "data", "raw")
processed_data_dir   <- file.path(project_repo_dir, "data", "processed")
out_dir          <- file.path(project_repo_dir, "outDir")

# FIX: verify required output directories exist before any downstream writes fail
# silently with an unhelpful error
for (d in c(processed_data_dir, out_dir)) {
  if (!dir.exists(d)) {
    stop("Required directory does not exist: ", d,
         "\nCreate it or update the path before re-running.")
  }
}
```

# Load Libraries

```{r}
# if (!requireNamespace("mogene10sttranscriptcluster.db", quietly = TRUE)) {
#   BiocManager::install("mogene10sttranscriptcluster.db", ask = FALSE)
# }
library(mogene10sttranscriptcluster.db)
library(AnnotationDbi)
```

# Load expression data

```{r}
expr_matrix <- readRDS(file.path(processed_data_dir, "expression_matrix_rma.rds"))

# Confirm the object is a numeric matrix (not a data frame or other class)
stopifnot(is.matrix(expr_matrix), is.numeric(expr_matrix))

cat("Loaded expression matrix with", nrow(expr_matrix), "probes and",
    ncol(expr_matrix), "samples\n")
```

# Extract probe IDs

```{r}
probe_ids <- rownames(expr_matrix)

if (is.null(probe_ids)) stop("Expression matrix has no rownames — probe IDs are missing.")

cat("Number of probes to annotate:", length(probe_ids), "\n")
```

# Map to gene annotations

```{r}
# Fetch annotations mapping probe IDs to gene-level identifiers.
# AnnotationDbi::select() may warn about 1-to-many mappings (one probe → many
# Ensembl IDs); this is expected and handled below.
annotations <- AnnotationDbi::select(
  mogene10sttranscriptcluster.db,
  keys    = probe_ids,
  columns = c("SYMBOL", "GENENAME", "ENSEMBL"),
  keytype = "PROBEID"
)

# --- Mapping summary ---
n_probes_input   <- length(probe_ids)
n_probes_mapped  <- length(unique(annotations$PROBEID[!is.na(annotations$SYMBOL)]))
n_probes_ensembl <- length(unique(annotations$PROBEID[!is.na(annotations$ENSEMBL)]))

cat("Probes submitted for annotation:           ", n_probes_input,   "\n")
cat("Probes mapped to a gene symbol:            ", n_probes_mapped,  "\n")
cat("Probes mapped to an Ensembl ID:            ", n_probes_ensembl, "\n")
cat("% probes WITHOUT a symbol:                 ",
    round(100 * (1 - n_probes_mapped / n_probes_input), 1), "%\n")
cat("% probes WITHOUT an Ensembl ID:            ",
    round(100 * (1 - n_probes_ensembl / n_probes_input), 1), "%\n")

# Preview 10 rows that have an Ensembl ID
annotations[!is.na(annotations$ENSEMBL), ][1:10, ]
```

# Resolve 1-to-many probe → gene mappings

```{r}
# Some probes map to multiple Ensembl IDs (isoform-level records); after
# filtering to non-NA SYMBOL rows we need a single representative row per
# PROBEID before merging.
#
# Strategy (intentional, not arbitrary):
#   1. Remove rows with no SYMBOL (unannotated probes).
#   2. For probes that have ANY Ensembl-mapped row, prefer that row over
#      Ensembl-NA rows — this makes the surviving annotation maximally
#      informative.
#   3. Among remaining ties (e.g. multiple Ensembl IDs for one probe), keep
#      the first — they share the same SYMBOL and GENENAME so choice is
#      inconsequential for the collapse step.
#
# This ensures reproducibility regardless of database row order.

annotations_clean <- annotations[!is.na(annotations$SYMBOL), ]

# FIX: prefer Ensembl-mapped rows when available (was arbitrary before)
# Order so Ensembl-present rows sort to the top within each PROBEID group
annotations_clean <- annotations_clean[
  order(annotations_clean$PROBEID,
        is.na(annotations_clean$ENSEMBL)), # FALSE (has Ensembl) sorts before TRUE
]

annotations_unique <- annotations_clean[!duplicated(annotations_clean$PROBEID), ]

cat("Unique annotated probes after deduplication:", nrow(annotations_unique), "\n")
```

# Merge with expression matrix

```{r}
# FIX: use all.x = TRUE (left join on annotations) rather than all.y = TRUE.
# all.y = TRUE kept unannotated probes in the output as NA-SYMBOL rows, which
# then collapsed to a single spurious "NA gene" row in the gene-level matrix.
# all.x = TRUE restricts the output strictly to probes that have a SYMBOL.
expr_annotated <- merge(
  annotations_unique,
  expr_matrix,
  by.x  = "PROBEID",
  by.y  = "row.names",
  all.x = TRUE   # ← FIXED (was all.y = TRUE)
)

# Sanity check: no NA symbols should survive the merge
n_na_symbol <- sum(is.na(expr_annotated$SYMBOL))
if (n_na_symbol > 0) {
  warning(n_na_symbol, " rows with NA SYMBOL remain after merge — investigate.")
} else {
  cat("All merged rows have a gene symbol — merge looks clean.\n")
}

# Preview
cat("\nRows with an Ensembl ID:\n")
print(head(expr_annotated[!is.na(expr_annotated$ENSEMBL), ], 10))
```

# Save probe-level annotated matrix

```{r}
# Save the full probe-level annotated matrix before collapsing.
# Preserves the ability to return to probe-level data if needed downstream.
write.csv(
  expr_annotated,
  file      = file.path(processed_data_dir, "expression_matrix_rma_annotated.csv"),
  row.names = FALSE,
  quote     = FALSE
)

saveRDS(
  expr_annotated,
  file = file.path(processed_data_dir, "expression_matrix_rma_annotated.rds")
)

cat("Probe-level annotated matrix saved:", nrow(expr_annotated), "probes\n")
```

# Collapse probes to gene level (highest mean expression per gene)

```{r}
# Identify expression columns positively (numeric columns only) rather than
# relying on a fixed list of annotation column names — more robust if the
# annotation package ever returns additional columns.
#
# FIX: detect expression columns by class rather than by name exclusion.
annot_cols <- c("PROBEID", "SYMBOL", "GENENAME", "ENSEMBL")
expr_cols  <- colnames(expr_annotated)[
  sapply(expr_annotated, is.numeric) & !colnames(expr_annotated) %in% annot_cols
]

cat("Expression columns detected:", length(expr_cols), "\n")

if (length(expr_cols) == 0) stop("No numeric expression columns found — check merge output.")

# Compute mean expression across all samples for each probe
expr_annotated$mean_expr <- rowMeans(expr_annotated[, expr_cols, drop = FALSE],
                                     na.rm = TRUE)

# For each gene symbol, retain only the probe with the highest mean expression.
# Order by descending mean expression so the first occurrence per SYMBOL is the
# best probe, then deduplicate.
expr_annotated_ordered <- expr_annotated[
  order(expr_annotated$SYMBOL, -expr_annotated$mean_expr), ]

expr_gene_level <- expr_annotated_ordered[!duplicated(expr_annotated_ordered$SYMBOL), ]

# Drop helper column now that collapsing is done
expr_gene_level$mean_expr <- NULL

cat("Probes before collapsing: ", nrow(expr_annotated), "\n")
cat("Genes after collapsing:   ", nrow(expr_gene_level), "\n")
```

# Inspect gene-level matrix

```{r}
cat("Gene-level matrix dimensions:", nrow(expr_gene_level), "genes x",
    ncol(expr_gene_level), "columns\n")

# FIX: was hardcoded expr_cols[1:4] — will now error if fewer than 4 samples.
# Preview up to 4 expression columns, however many exist.
preview_cols <- head(expr_cols, min(4, length(expr_cols)))
print(head(expr_gene_level[, c("PROBEID", "SYMBOL", "GENENAME", preview_cols)], 20))
```

# Save gene-level matrix

```{r}
write.csv(
  expr_gene_level,
  file      = file.path(processed_data_dir, "expression_matrix_rma_gene_level.csv"),
  row.names = FALSE,
  quote     = FALSE
)

saveRDS(
  expr_gene_level,
  file = file.path(processed_data_dir, "expression_matrix_rma_gene_level.rds")
)

cat("Gene-level matrix saved:", nrow(expr_gene_level), "genes\n")
cat("Annotation complete.\n")
```

# Session info for reproducibility

```{r}
writeLines(
  capture.output(sessionInfo()),
  file.path(out_dir, "sessionInfo_annotation.txt")
)
```
