---
title: "probe_to_gene_mapping"
author: "Melanie Smith"
date: "2026-05-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Clean-up environment

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc()                            # free up memory and report the memory usage
options(
  max.print        = .Machine$integer.max,
  scipen           = 999,
  stringsAsFactors = FALSE,
  dplyr.summarise.inform = FALSE
)
set.seed(42) # set seed for reproducibility
```

# Set project directories and file paths

```{r}
project_name  <- "20260421_natashaHarvey_gata2_mouse"
base_repo_dir <- "/home/melanie-smith/workDir/natashaHarvey"

project_repo_dir <- file.path(base_repo_dir, project_name)

processed_data_dir <- file.path(project_repo_dir, "data", "processed")
out_dir            <- file.path(project_repo_dir, "outDir")

# Verify required directories exist before any downstream writes fail silently
for (d in c(processed_data_dir, out_dir)) {
  if (!dir.exists(d)) {
    stop("Required directory does not exist: ", d,
         "\nCreate it or update the path before re-running.")
  }
}
```

# Load libraries

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

# mogene10sttranscriptcluster.db may need to be installed manually if not present:
# BiocManager::install("mogene10sttranscriptcluster.db")
if (!requireNamespace("mogene10sttranscriptcluster.db", quietly = TRUE)) {
  BiocManager::install("mogene10sttranscriptcluster.db", ask = FALSE, update = FALSE)
}

required_pkgs <- c("Biobase", "AnnotationDbi", "mogene10sttranscriptcluster.db", "dplyr")

for (pkg in required_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    BiocManager::install(pkg, ask = FALSE, update = FALSE)
  }
  library(pkg, character.only = TRUE)
}
```

# Load filtered ExpressionSet

```{r}
# Load the intensity-filtered, RMA-normalised ExpressionSet produced by the
# filtering step (norm_data_filtered.rds). Using the ExpressionSet object
# directly — rather than extracting a plain matrix — keeps expression data,
# sample metadata (pData), and probe annotations (fData) together in one
# validated object throughout this script.
norm_filtered <- readRDS(file.path(processed_data_dir, "norm_data_filtered.rds"))

# Validate the object is internally consistent before proceeding
validObject(norm_filtered)

cat("Loaded filtered ExpressionSet:\n")
cat("  Probes :", nrow(norm_filtered), "\n")
cat("  Samples:", ncol(norm_filtered), "\n")
cat("\nSample names:\n")
print(sampleNames(norm_filtered))
cat("\npData:\n")
print(pData(norm_filtered))
```

# Fetch probe annotations

```{r}
# Query the MoGene-1.0-ST annotation package for gene-level identifiers.
# featureNames() retrieves probe IDs directly from the ExpressionSet.
# ENSEMBL is included alongside SYMBOL and GENENAME for use in downstream
# pathway analyses (e.g. fgsea, g:Profiler) that require Ensembl IDs.
#
# Note: AnnotationDbi::select() will warn about 1-to-many mappings —
# this is expected and is handled explicitly in the steps below.
anno <- AnnotationDbi::select(
  mogene10sttranscriptcluster.db,
  keys    = featureNames(norm_filtered),
  columns = c("SYMBOL", "GENENAME", "ENSEMBL"),
  keytype = "PROBEID"
)

# --- Annotation mapping summary ---
n_probes_input   <- length(featureNames(norm_filtered))
n_probes_symbol  <- length(unique(anno$PROBEID[!is.na(anno$SYMBOL)]))
n_probes_ensembl <- length(unique(anno$PROBEID[!is.na(anno$ENSEMBL)]))

cat("Probes submitted for annotation:  ", n_probes_input,   "\n")
cat("Probes mapped to a gene symbol:   ", n_probes_symbol,  "\n")
cat("Probes mapped to an Ensembl ID:   ", n_probes_ensembl, "\n")
cat("% probes WITHOUT a symbol:        ",
    round(100 * (1 - n_probes_symbol  / n_probes_input), 1), "%\n")
cat("% probes WITHOUT an Ensembl ID:   ",
    round(100 * (1 - n_probes_ensembl / n_probes_input), 1), "%\n")
```

# Remove unannotated probes (no gene symbol)

```{r}
# Probes with no SYMBOL cannot be interpreted biologically and are excluded.
# This mirrors the published approach: subset(anno, !is.na(SYMBOL))
anno <- subset(anno, !is.na(SYMBOL))

# Report unique probes, not rows
cat("Rows after removing no-SYMBOL probes:          ", nrow(anno), "\n")
cat("Unique probes after removing no-SYMBOL probes: ", 
    length(unique(anno$PROBEID)), "\n")
```

# Identify and remove ambiguously mapped probes

```{r}
# Some probe IDs map to more than one gene symbol (ambiguous assignments).
# Rather than arbitrarily selecting one mapping, we exclude these probes
# entirely — consistent with the peer-reviewed workflow this pipeline follows.
#
# Step 1: count distinct gene symbols per probe
anno_grouped   <- dplyr::group_by(anno, PROBEID)
anno_summarised <- dplyr::summarise(anno_grouped,
                                    no_of_matches = n_distinct(SYMBOL))

# Step 2: identify probes with more than one symbol
anno_multi <- dplyr::filter(anno_summarised, no_of_matches > 1)

cat("Probes mapping to multiple gene symbols (to exclude):", nrow(anno_multi), "\n")

# Step 3: remove ambiguous probes from annotation table
anno <- dplyr::filter(anno, !PROBEID %in% anno_multi$PROBEID)

cat("Probes remaining after removing ambiguous mappings:  ", 
    length(unique(anno$PROBEID)), "\n")
```

# Resolve remaining 1-to-many probe → Ensembl mappings

```{r}
# After removing multi-symbol probes, some probes may still have multiple rows
# due to multiple Ensembl IDs for the same gene (isoform-level records).
# SYMBOL and GENENAME are identical across these rows — only ENSEMBL differs.
#
# Strategy: prefer rows that have an Ensembl ID (sort Ensembl-NA rows to
# the bottom), then keep the first row per PROBEID. This retains an Ensembl ID
# wherever one exists without discarding the probe entirely.
anno <- anno[order(anno$PROBEID, is.na(anno$ENSEMBL)), ]
anno <- anno[!duplicated(anno$PROBEID), ]

cat("Unique probes after Ensembl deduplication:", nrow(anno), "\n")
cat("\nPreview of annotation table:\n")
print(head(anno, 10))
```

# Exclude ambiguous probes from the ExpressionSet

```{r}
# Probes to exclude: those with no SYMBOL OR those with ambiguous mappings
ids_to_exclude <- featureNames(norm_filtered) %in% anno_multi$PROBEID |
                  !featureNames(norm_filtered) %in% anno$PROBEID

cat("Probe exclusion summary:\n")
print(table(ids_to_exclude))
# ids_to_exclude
# FALSE  TRUE 
# 19377  9991 

eset_final <- subset(norm_filtered, !ids_to_exclude)
validObject(eset_final)

cat("\nExpressionSet after probe exclusion:\n")
cat("  Probes :", nrow(eset_final), "\n")   # expect 19,377
cat("  Samples:", ncol(eset_final), "\n")
```

# Add annotations to featureData of the ExpressionSet

```{r}
# Store probe annotations directly in the featureData slot of the ExpressionSet.
# This keeps SYMBOL, GENENAME, and ENSEMBL travelling with the expression data
# as a single validated object — no separate annotation data frames needed
# downstream.
#
# Step 1: seed featureData with PROBEID (rownames of fData by default)
fData(eset_final)$PROBEID <- rownames(fData(eset_final))

# Step 2: left-join annotation columns into featureData.
# left_join preserves all rows of fData(eset_final) and adds SYMBOL,
# GENENAME, ENSEMBL from anno for matching PROBEIDs only.
fData(eset_final) <- dplyr::left_join(fData(eset_final), anno, by = "PROBEID")

# Step 3: restore rownames (left_join drops them)
rownames(fData(eset_final)) <- fData(eset_final)$PROBEID

# Validate the final annotated object
validObject(eset_final)

cat("featureData annotation added successfully.\n")
cat("\nPreview of fData:\n")
print(head(fData(eset_final), 10))

# Confirm no missing symbols remain in fData
n_na <- sum(is.na(fData(eset_final)$SYMBOL))
if (n_na > 0) {
  warning(n_na, " probes with NA SYMBOL remain in fData — investigate.")
} else {
  cat("\nAll probes in fData have a gene symbol — annotation looks clean.\n")
}
```

# Final ExpressionSet summary

```{r}
cat("=== Final annotated ExpressionSet summary ===\n")
cat("Probes (features): ", nrow(eset_final), "\n")
cat("Samples:           ", ncol(eset_final), "\n")
cat("\nfeatureData columns:", paste(colnames(fData(eset_final)), collapse = ", "), "\n")
cat("phenoData columns:  ", paste(colnames(pData(eset_final)), collapse = ", "), "\n")

cat("\nSample group assignments:\n")
print(pData(eset_final)[, c("sample_name", "group")])

cat("\nfData preview (first 10 probes):\n")
print(head(fData(eset_final)[, c("PROBEID", "SYMBOL", "GENENAME", "ENSEMBL")], 10))
```

# Save annotated ExpressionSet

```{r}
# Save the fully annotated ExpressionSet as the primary output of this script.
# Downstream scripts (DE analysis, visualisation) load this single object
# and have immediate access to both expression values and annotations.
saveRDS(
  eset_final,
  file = file.path(processed_data_dir, "eset_annotated_filtered.rds")
)

# Also save a human-readable CSV of fData for reference
write.csv(
  fData(eset_final)[, c("PROBEID", "SYMBOL", "GENENAME", "ENSEMBL")],
  file      = file.path(out_dir, "probe_annotation_table.csv"),
  row.names = FALSE,
  quote     = FALSE
)

cat("Annotated ExpressionSet saved to: ",
    file.path(processed_data_dir, "eset_annotated_filtered.rds"), "\n")
cat("Annotation table saved to:        ",
    file.path(out_dir, "probe_annotation_table.csv"), "\n")
```

# Session info for reproducibility

```{r}
writeLines(
  capture.output(sessionInfo()),
  file.path(out_dir, "sessionInfo_annotation.txt")
)

cat("Annotation script complete.\n")
```
