---
title: "lmFit_gata2_vs_con"
author: "Melanie Smith"
date: "2026-04-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Clean-up environment

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max,
        scipen = 999,
        stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE) # Print everything without truncation, avoid scientific notation, keep strings as characters, silence dplyr's "grouped output" informational messages
set.seed(42) # set seed for reproducibility
```

# Set project directories and file paths

```{r}
project_name <- "20260421_natashaHarvey_gata2_mouse"
base_repo_dir <- "/home/melanie-smith/workDir/natashaHarvey"

project_repo_dir <- file.path(base_repo_dir, project_name)

processed_data <- file.path(project_repo_dir, "data", "processed")
out_dir <- file.path(project_repo_dir, "outDir")

dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```

# Load libraries

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

required_pkgs <- c("limma")

for (pkg in required_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    BiocManager::install(pkg, ask = FALSE, update = FALSE)
  }
  library(pkg, character.only = TRUE)
}
```

# Load gene-level expression matrix

```{r}
# Load the gene-level matrix produced by script 02 (one row per gene,
# best probe selected by highest mean expression across all samples)
expr_annotated <- readRDS(file.path(processed_data, "expression_matrix_rma_gene_level.rds"))

cat("Loaded gene-level expression matrix with", nrow(expr_annotated), "genes\n")
```

# Separate annotation columns from expression values

```{r}
# Annotation columns produced by script 02
annot_cols <- c("PROBEID", "SYMBOL", "GENENAME", "ENSEMBL")

# Grab the sample IDs (i.e. the CEL file sample columns)
expr_cols <- setdiff(colnames(expr_annotated), annot_cols)

# Split into annotation and numeric expression data frames
annot_df <- expr_annotated[, annot_cols, drop = FALSE]
expr_mat  <- as.matrix(expr_annotated[, expr_cols])
rownames(expr_mat) <- expr_annotated$PROBEID

cat("Annotation columns:", paste(annot_cols, collapse = ", "), "\n")
cat("Expression columns (", length(expr_cols), " samples):\n", sep = "")
print(expr_cols)
```

# Define sample groups from CEL filenames

```{r}
# Group labels are encoded in filenames as the second '_'-delimited field:
#   JK_con_A_(MoGene-1_0-st-v1).CEL   -> "con"
#   JK_GATA2_A_(MoGene-1_0-st-v1).CEL -> "GATA2"
group_labels <- sub("^JK_([^_]+)_.*$", "\\1", expr_cols)

# Coerce to factor with control as the reference level
group <- factor(group_labels, levels = c("con", "GATA2"))
# sanity check
cat("Sample group assignments:\n")
print(data.frame(sample = expr_cols, group = group))
```

# Build design matrix and fit linear model (limma)

```{r}
# One-way design: ~0 + group gives explicit group means (no intercept),
# which makes the contrast definition straightforward
design <- model.matrix(~0 + group)
colnames(design) <- levels(group)

cat("Design matrix:\n")
print(design)

fit <- lmFit(expr_mat, design)
```

# Define contrast: GATA2 knockdown vs control

```{r}
contrast_matrix <- makeContrasts(
  GATA2_vs_con = GATA2 - con,
  levels = design
)

fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2) # empirical Bayes moderation (produces moderated F-statistics)
```

# Extract full results table

```{r}
# Extract all probes, unsorted, no filtering
results <- topTable(
  fit2,
  coef      = "GATA2_vs_con",
  number     = Inf,
  sort.by    = "none",
  adjust.method = "BH" # Benjamini-Hochberg FDR correction (stepup p-value)
  )

cat("Results table dimensions:", nrow(results), "rows x", ncol(results), "columns\n")
```

# Merge results with probe annotations

```{r}
# Bring annotation columns back in alongside the limma statistics
results$PROBEID <- rownames(results)

results_annotated <- merge(
  annot_df,
  results,
  by = "PROBEID",
  all.y = TRUE
)

# Compute fold-change direction label to match old "Fold-Change (Description)" column
results_annotated$FC_description <- ifelse(
  results_annotated$logFC > 0, "Upregulated in GATA2",
  ifelse(results_annotated$logFC < 0, "Downregulated in GATA2", "No change")
)

cat("Annotated results table dimensions:", nrow(results_annotated), "rows x",
    ncol(results_annotated), "columns\n")
```

# Rename columns to match legacy TAC-style output

```{r}
# Annotation columns produced by script 02
results_final <- results_annotated |>
  dplyr::select(
    `Transcript Cluster ID` = PROBEID,
    `Gene Assignment`        = GENENAME,
    `Gene Symbol`            = SYMBOL,
    `mRNA Assignment`        = ENSEMBL,
    `p-value(GATA2 vs. con)` = P.Value,
    `stepup(p-value(GATA2 vs. con))` = adj.P.Val,
    `logFC(GATA2 vs. con)`   = logFC,
    `log-odds(B)`            = B,
    `Fold-Change (Description)` = FC_description,
    `AveExpr`                = AveExpr
  )

cat("Final column names:\n")
print(colnames(results_final))
```

# Save full unfiltered results

```{r}
write.csv(
  results_final,
  file = file.path(out_dir, "lmFit_GATA2_vs_con_full.csv"),
  row.names = FALSE,
  quote = FALSE
)

saveRDS(
  results_final,
  file = file.path(out_dir, "lmFit_GATA2_vs_con_full.rds")
)

cat("Results saved to:", file.path(out_dir, "lmFit_GATA2_vs_con_full.csv"), "\n")
```

# Quick summary of results

```{r}
cat("--- Results summary ---\n")
cat("Total probes tested:", nrow(results_final), "\n")
cat("Probes with FDR < 0.05:",
    sum(results_final$`stepup(p-value(GATA2 vs. con))` < 0.05, na.rm = TRUE), "\n")
cat("Probes with FDR < 0.01:",
    sum(results_final$`stepup(p-value(GATA2 vs. con))` < 0.01, na.rm = TRUE), "\n")
cat("Upregulated in GATA2 (logFC > 0, FDR < 0.05):",
    sum(results_final$`logFC(GATA2 vs. con)` > 0 &
          results_final$`stepup(p-value(GATA2 vs. con))` < 0.05, na.rm = TRUE), "\n")
cat("Downregulated in GATA2 (logFC < 0, FDR < 0.05):",
    sum(results_final$`logFC(GATA2 vs. con)` < 0 &
          results_final$`stepup(p-value(GATA2 vs. con))` < 0.05, na.rm = TRUE), "\n")
```

# Session info for reproducibility

```{r}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo_lmFit.txt"))

cat("DE complete.\n")
```
