---
title: "lmFit_gata2_vs_con"
author: "Melanie Smith"
date: "2026-04-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Clean-up environment

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max,
        scipen = 999,
        stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE) # Print everything without truncation, avoid scientific notation, keep strings as characters, silence dplyr's "grouped output" informational messages
set.seed(42) # set seed for reproducibility
```

# Set project directories and file paths

```{r}
project_name <- "20260421_natashaHarvey_gata2_mouse"
base_repo_dir <- "/home/melanie-smith/workDir/natashaHarvey"

project_repo_dir <- file.path(base_repo_dir, project_name)

processed_data <- file.path(project_repo_dir, "data", "processed")
out_dir <- file.path(project_repo_dir, "outDir")

dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```

# Load libraries

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

required_pkgs <- c("limma", "tidyverse", "ggplot2", "ggrepel", "Biobase")

for (pkg in required_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    BiocManager::install(pkg, ask = FALSE, update = FALSE)
  }
  library(pkg, character.only = TRUE)
}
```

# Load gene-level expression matrix

```{r}
# Load the annotated, filtered ExpressionSet produced by script 02
eset <- readRDS(file.path(processed_data, "eset_annotated_filtered.rds"))

# Validate before proceeding
validObject(eset)

cat("Loaded annotated ExpressionSet:\n")
cat("  Probes :", nrow(eset), "\n")
cat("  Samples:", ncol(eset), "\n")
```

# Separate annotation columns from expression values

```{r}
# Extract expression matrix and annotations from the ExpressionSet slots
# Numeric expression matrix — probes x samples
expr_mat <- exprs(eset)

# Annotation data frame from featureData slot
annot_df <- fData(eset)[, c("PROBEID", "SYMBOL", "GENENAME", "ENSEMBL"), drop = FALSE]

# Sample column names for reference
expr_cols <- colnames(expr_mat)

cat("Probes:  ", nrow(expr_mat), "\n")
cat("Samples: ", length(expr_cols), "\n")
cat("Annotation columns:", paste(colnames(annot_df), collapse = ", "), "\n")
cat("Sample columns:\n")
print(expr_cols)

# Sanity check — rownames of expr_mat should match PROBEID in annot_df
stopifnot(
  "Probe ID mismatch between exprs() and fData() — check ExpressionSet integrity" =
    all(rownames(expr_mat) == annot_df$PROBEID)
)
```

# Define sample groups from CEL filenames

```{r}
# Sample names are already cleaned to "group_replicate" format (e.g. "con_A", "GATA2_A")
# Strip the trailing _A/_B/_C/_D to recover the group label
group_labels <- sub("_[A-Za-z]$", "", expr_cols)

# Coerce to factor with control as the reference level
group <- factor(group_labels, levels = c("con", "GATA2"))

# Fail early if any unexpected group labels are produced
if (any(is.na(group))) {
  stop("Unexpected group labels detected — check sample name format in expr_cols.")
}

# Sanity check
cat("Sample group assignments:\n")
print(data.frame(sample = expr_cols, group = group))
```

# Build design matrix and fit linear model (limma)

```{r}
# One-way design: ~0 + group gives explicit group means (no intercept),
# which makes the contrast definition straightforward
design <- model.matrix(~0 + group)
colnames(design) <- levels(group)

cat("Design matrix:\n")
print(design)

fit <- lmFit(expr_mat, design)
```

# Define contrast: GATA2 knockdown vs control

```{r}
contrast_matrix <- makeContrasts(
  GATA2_vs_con = GATA2 - con,
  levels = design
)

fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2, 
               robust = TRUE, # For added protection against outlier probes influencing the empirical Bayes variance estimation
               trend = TRUE) # empirical Bayes moderation (produces moderated t- and F-statistics)
```

# Extract full results table

```{r}
# Extract all probes, unsorted, no filtering
results <- topTable(
  fit2,
  coef      = "GATA2_vs_con",
  number     = Inf,
  sort.by    = "none",
  adjust.method = "BH" # Benjamini-Hochberg FDR correction (stepup p-value)
  )

cat("Results table dimensions:", nrow(results), "rows x", ncol(results), "columns\n")
```

# Merge results with probe annotations

```{r}
# Bring annotation columns back in alongside the limma statistics
results$PROBEID <- rownames(results)

results_annotated <- merge(
  annot_df,
  results,
  by    = "PROBEID",
  all.x = TRUE   # left join on annotations — drops any result rows with no annotation
)

# make sure no rows were dropped in the merge
stopifnot(
  "Row count changed after merge — check for PROBEID mismatches" =
    nrow(results_annotated) == nrow(annot_df)
)

# Compute fold-change direction label to match old "Fold-Change (Description)" column
results_annotated$FC_description <- ifelse(
  results_annotated$logFC > 0, "Upregulated in GATA2",
  ifelse(results_annotated$logFC < 0, "Downregulated in GATA2", "No change")
)

cat("Annotated results table dimensions:", nrow(results_annotated), "rows x",
    ncol(results_annotated), "columns\n")
```

# Save full unfiltered results

```{r}

write.table(
  results_annotated,
  file = file.path(out_dir, "lmFit_GATA2_vs_con_full.txt"),
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)

saveRDS(
  results_annotated,
  file = file.path(processed_data, "lmFit_GATA2_vs_con_full.rds")
  )

# Fix
cat("Results saved — TXT:", file.path(out_dir, "lmFit_GATA2_vs_con_full.txt"), "\n")
cat("Results saved — RDS:", file.path(processed_data, "lmFit_GATA2_vs_con_full.rds"), "\n")
```

# Quick summary of results

```{r}
cat("--- Results summary ---\n")
cat("Total probes tested:", nrow(results_annotated), "\n")
cat("Probes with FDR < 0.05:",
    sum(results_annotated$adj.P.Val < 0.05, na.rm = TRUE), "\n")
cat("Probes with FDR < 0.01:",
    sum(results_annotated$adj.P.Val < 0.01, na.rm = TRUE), "\n")
cat("Upregulated in GATA2 (logFC > 0, FDR < 0.05):",
    sum(results_annotated$logFC > 0 &
          results_annotated$adj.P.Val < 0.05, na.rm = TRUE), "\n")
cat("Downregulated in GATA2 (logFC < 0, FDR < 0.05):",
    sum(results_annotated$logFC < 0 &
          results_annotated$adj.P.Val < 0.05, na.rm = TRUE), "\n")
cat("Sig. & |logFC| > 1 (FDR < 0.05):",
    sum(abs(results_annotated$logFC) > 1 &
          results_annotated$adj.P.Val < 0.05, na.rm = TRUE), "\n")
```

# Prepare plot data — classify probes for highlighting

```{r}
# Significance thresholds
fdr_thresh <- 0.05
lfc_thresh <- 1

# Classify each probe into one of three groups for colour coding
results_annotated$sig_group <- dplyr::case_when(
  results_annotated$adj.P.Val < fdr_thresh & results_annotated$logFC >  lfc_thresh ~ "Up",
  results_annotated$adj.P.Val < fdr_thresh & results_annotated$logFC < -lfc_thresh ~ "Down",
  TRUE                                                                               ~ "NS"
)
results_annotated$sig_group <- factor(results_annotated$sig_group,
                                      levels = c("Up", "Down", "NS"))

# Count each group for legend labels
n_up   <- sum(results_annotated$sig_group == "Up",   na.rm = TRUE)
n_down <- sum(results_annotated$sig_group == "Down", na.rm = TRUE)
n_ns   <- sum(results_annotated$sig_group == "NS",   na.rm = TRUE)

cat("Up:            ", n_up,   "\n")
cat("Down:          ", n_down, "\n")
cat("Not significant:", n_ns,   "\n")

# Shared colour scale
sig_colours <- c("Up"   = "#E74C3C",
                 "Down" = "#2a16db",
                 "NS"   = "grey75")

sig_labels <- c(
  "Up"   = paste0("Up in GATA2 (n=", n_up,   ")"),
  "Down" = paste0("Down in GATA2 (n=", n_down, ")"),
  "NS"   = paste0("Not significant (n=", n_ns, ")")
)

# Top probes to label: top 10 up and top 10 down by adj.P.Val among significant hits
top_up <- results_annotated |>
  dplyr::filter(sig_group == "Up") |>
  dplyr::slice_min(order_by = adj.P.Val, n = 6)

top_down <- results_annotated |>
  dplyr::filter(sig_group == "Down") |>
  dplyr::slice_min(order_by = adj.P.Val, n = 6)

top_probes <- rbind(top_up, top_down)
```

# MA plot

```{r ma_plot, fig.width = 8, fig.height = 6, dpi = 300}

ma_plot <- ggplot(results_annotated,
                  aes(x      = AveExpr,
                      y      = logFC,
                      colour = sig_group)) +

  # NS points first so coloured points render on top
  geom_point(data  = dplyr::filter(results_annotated, sig_group == "NS"),
             size  = 0.8, alpha = 0.35) +
  geom_point(data  = dplyr::filter(results_annotated, sig_group != "NS"),
             size  = 1.2, alpha = 0.85) +

  # Reference lines
  geom_hline(yintercept = 0,
             colour = "grey30", linewidth = 0.5) +
  geom_hline(yintercept = c(-lfc_thresh, lfc_thresh),
             linetype = "dashed", colour = "grey40", linewidth = 0.4) +

  # Gene symbol labels for top hits
  ggrepel::geom_text_repel(
    data           = top_probes,
    aes(label      = SYMBOL),
    size           = 2.8,
    fontface       = "italic",
    show.legend    = FALSE,
    max.overlaps   = Inf,
    box.padding    = 0.4,
    point.padding  = 0.3,
    segment.colour = "grey50",
    segment.size   = 0.3
  ) +

  scale_colour_manual(
    name   = NULL,
    values = sig_colours,
    labels = sig_labels
  ) +

  labs(
    title    = "MA plot — GATA2 vs control",
    subtitle = paste0("FDR < ", fdr_thresh, " and |log\u2082FC| > ", lfc_thresh,
                      " highlighted (", n_up + n_down, " probes)"),
    x        = "Average log\u2082 expression (AveExpr)",
    y        = expression(log[2]~"fold-change (GATA2 / control)")
  ) +

  theme_bw(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(colour = "grey40", size = 11),
    legend.position  = "bottom",
    legend.text      = element_text(size = 10),
    panel.grid.minor = element_blank()
  )

print(ma_plot)

# Save
png(file.path(out_dir, "MA_plot_GATA2_vs_con.png"),
    width = 8, height = 6, units = "in", res = 300)
print(ma_plot)
dev.off()

cat("MA plot saved to:", file.path(out_dir, "MA_plot_GATA2_vs_con.png"), "\n")
```

# Volcano plot

```{r volcano_plot, fig.width = 8, fig.height = 7, dpi = 300}
volcano_plot <- ggplot(results_annotated,
       aes(x      = logFC,
           y      = -log10(adj.P.Val),
           colour = sig_group)
       ) +
  
  scale_x_continuous(
    breaks = scales::breaks_width(1)
  ) +

  # NS points first so coloured points render on top
  geom_point(data  = dplyr::filter(results_annotated, sig_group == "NS"),
             size  = 0.8, alpha = 0.35) +
  geom_point(data  = dplyr::filter(results_annotated, sig_group != "NS"),
             size  = 1.2, alpha = 0.85) +
  
  # Gene symbol labels for top hits - up only (red)
  ggrepel::geom_text_repel(
    data           = dplyr::filter(top_probes, sig_group == "Up"),
    aes(label      = SYMBOL),
    size           = 4,
    fontface       = "italic",
    show.legend    = FALSE,
    max.overlaps   = Inf,
    force          = 3,
    box.padding    = 0.7,
    point.padding  = 0.4,
    segment.colour = "black",
    segment.size   = 0.3,
    min.segment.length = 0
  ) +

  # Gene symbol labels for top hits - Down only (blue)
  ggrepel::geom_text_repel(
    data           = dplyr::filter(top_probes, sig_group == "Down"),
    aes(label      = SYMBOL),
    size           = 4,
    fontface       = "italic",
    show.legend    = FALSE,
    max.overlaps   = Inf,
    force          = 3,
    box.padding    = 0.47,
    point.padding  = 0.3,
    segment.colour = "black",
    segment.size   = 0.3,
    min.segment.length = 0
  ) +
  
  scale_colour_manual(
    name   = NULL,
    values = sig_colours,
    labels = sig_labels
  ) +
  
  labs(
    x        = expression(log[2]~"fold-change (GATA2 / control)"),
    y        = expression(-log[10]~"(FDR-adjusted"~italic(p)*"-value)")
  ) +
  
theme_classic(base_size = 13, base_family = "Helvetica") +
  theme(
    # All text black and Helvetica
    text             = element_text(family = "Helvetica", colour = "black"),
    axis.text        = element_text(family = "Helvetica", colour = "black"),
    
    # Remove colour legend
    legend.position  = "none",
    
    # Solid black x- and y-axis lines
    axis.line        = element_line(colour = "black", linewidth = 0.5),
    
    # Remove all grid lines and panel border
    panel.grid       = element_blank(),
    panel.border     = element_blank()
  )

print(volcano_plot)

svg(file.path(out_dir, "volcano_plot_GATA2_vs_con.svg"),
    width  = 4.76,
    height = 5)
print(volcano_plot)
dev.off()
cat("Volcano plot saved to:", file.path(out_dir, "volcano_plot_GATA2_vs_con.png"), "\n")
```

# Gene-level probe summary table

```{r}
# For every gene SYMBOL that has at least one significant probe,
# summarise across all probes mapping to that gene.
# Concordance is defined as all significant probes pointing in the same direction.

gene_summary <- results_annotated %>%
  dplyr::group_by(SYMBOL, GENENAME) %>%
  dplyr::summarise(
    n_probes        = dplyr::n(),
    n_sig           = sum(adj.P.Val < fdr_thresh & abs(logFC) > lfc_thresh,
                          na.rm = TRUE),
    pct_sig         = round(100 * n_sig / n_probes, 1),
    mean_logFC_sig  = ifelse(
      n_sig > 0,
      round(mean(logFC[adj.P.Val < fdr_thresh & abs(logFC) > lfc_thresh],
                 na.rm = TRUE), 3),
      NA_real_
    ),
    concordant      = ifelse(
      n_sig > 1,
      # All significant probes have the same sign of logFC
      length(unique(sign(logFC[adj.P.Val < fdr_thresh &
                                 abs(logFC) > lfc_thresh]))) == 1,
      # Only one significant probe — concordance not applicable
      NA
    ),
    direction       = dplyr::case_when(
      n_sig == 0              ~ "None significant",
      is.na(concordant)       ~ ifelse(mean_logFC_sig > 0,
                                       "Up (single probe)",
                                       "Down (single probe)"),
      concordant & mean_logFC_sig > 0 ~ "Concordant Up",
      concordant & mean_logFC_sig < 0 ~ "Concordant Down",
      !concordant                     ~ "Discordant"
    ),
    .groups = "drop"
  ) |>
  dplyr::filter(n_sig > 0) |>
  dplyr::arrange(dplyr::desc(pct_sig), dplyr::desc(n_sig))

cat("Genes with at least one significant probe:", nrow(gene_summary), "\n")
cat("Genes with all probes significant (100%):",
    sum(gene_summary$pct_sig == 100), "\n")
cat("Genes with discordant probe directions:",
    sum(gene_summary$direction == "Discordant", na.rm = TRUE), "\n\n")

print(head(gene_summary, 20))

# Save
write.csv(
  gene_summary,
  file      = file.path(out_dir, "gene_level_probe_summary.csv"),
  row.names = FALSE,
  quote     = FALSE
)

cat("Gene-level summary saved to:",
    file.path(out_dir, "gene_level_probe_summary.csv"), "\n")
```

# Session info for reproducibility

```{r}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo_lmFit.txt"))

cat("DE complete.\n")
```
