---
title: "compare_anova_results_new_vs_old"
author: "Melanie Smith"
date: "2026-04-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Clean-up environment

```{r}
rm(list = ls(all.names = TRUE)) # will clear all objects including hidden objects
gc() # free up memory and report the memory usage
options(max.print = .Machine$integer.max,
        scipen = 999,
        stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE) # Print everything without truncation, avoid scientific notation, keep strings as characters, silence dplyr's "grouped output" informational messages
set.seed(42) # set seed for reproducibility
```

# Set project directories and file paths

```{r}
project_name <- "20260421_natashaHarvey_gata2_mouse"
base_repo_dir <- "/home/melanie-smith/workDir/natashaHarvey"

project_repo_dir <- file.path(base_repo_dir, project_name)

processed_data <- file.path(project_repo_dir, "data", "processed")
out_dir        <- file.path(project_repo_dir, "outDir")
downloads_dir  <- file.path(project_repo_dir, "downloads", "Jan_ANOVA_con_vs_GATA2_mouse")
```

# Load libraries

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

required_pkgs <- c("readxl", "dplyr", "ggplot2")

for (pkg in required_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
  library(pkg, character.only = TRUE)
}
```

# Load new ANOVA results (script 03 output)

```{r}
results_new <- readRDS(file.path(out_dir, "ANOVA_GATA2_vs_con_full.rds"))

cat("New results dimensions:", nrow(results_new), "rows x", ncol(results_new), "cols\n")
cat("New results column names:\n")
print(colnames(results_new))
```

# Load old ANOVA results (.xls)

```{r}
old_xls_path <- file.path(downloads_dir, "Jan ANOVA con vs GATA2 mouse.xls")

results_old <- readxl::read_xls(old_xls_path)

cat("Old results dimensions:", nrow(results_old), "rows x", ncol(results_old), "cols\n")
cat("Old results column names:\n")
print(colnames(results_old))
```

# Inspect old results table

```{r}
# Preview first few rows to understand structure and confirm key column content
head(results_old, 10)
```

# Standardise join key (Transcript Cluster ID / PROBEID)

```{r}
# Coerce both ID columns to character to avoid any type mismatch on join
results_new <- results_new |>
  dplyr::rename(probe_id = `Transcript Cluster ID`) |>
  dplyr::mutate(probe_id = as.character(probe_id))

# Identify the Transcript Cluster ID column in the old table
# (column name may differ slightly from the new table)
old_id_col <- "Transcript Cluster ID"
stopifnot("Expected ID column not found in old results" = old_id_col %in% colnames(results_old))

results_old <- results_old |>
  dplyr::rename(probe_id = all_of(old_id_col)) |>
  dplyr::mutate(probe_id = as.character(probe_id))
```

# Check probe ID overlap

```{r}
ids_new <- results_new$probe_id
ids_old <- results_old$probe_id

n_new        <- length(ids_new)
n_old        <- length(ids_old)
n_shared     <- length(intersect(ids_new, ids_old))
n_only_new   <- length(setdiff(ids_new, ids_old))
n_only_old   <- length(setdiff(ids_old, ids_new))

cat("---------------------------------------\n")
cat("Probe ID overlap summary\n")
cat("---------------------------------------\n")
cat("Probes in new results only:          ", n_new, "\n")
cat("Probes in old results only:          ", n_old, "\n")
cat("Probes shared (intersection):        ", n_shared, "\n")
cat("Probes unique to new results:        ", n_only_new, "\n")
cat("Probes unique to old results:        ", n_only_old, "\n")
cat("Overlap as % of new results:         ",
    round(n_shared / n_new * 100, 1), "%\n")
cat("Overlap as % of old results:         ",
    round(n_shared / n_old * 100, 1), "%\n")
cat("---------------------------------------\n")
```

# Join tables on shared probe IDs

```{r}
# We use an inner join, retaining only the 21,312 probe IDs present in both tables.
# The ~6,958 probes unique to the old table are the redundant probes that were
# collapsed away in our pipeline (script 02) — their absence here is expected
# and does not represent data loss. The 13 probes unique to our new results are
# negligible annotation edge cases and are also excluded.
results_joined <- dplyr::inner_join(
  results_new |> dplyr::select(probe_id,
                                `Gene Symbol`,
                                logFC_new = `logFC(GATA2 vs. con)`,
                                fdr_new   = `stepup(p-value(GATA2 vs. con))`),
  results_old |> dplyr::select(probe_id,
                                ratio_old = `Ratio(GATA2 vs. con)`,
                                fdr_old   = `stepup(p-value(GATA2 vs. con))`),
  by = "probe_id"
)

cat("Joined table dimensions:", nrow(results_joined), "rows x",
    ncol(results_joined), "cols\n")
```


# Log2-transform the old ratio column

```{r}
# The old pipeline (TAC) reports fold-change as a linear ratio:
# values > 1 mean upregulated in GATA2 (e.g. 2.0 = twice as expressed),
# values < 1 mean downregulated (e.g. 0.5 = half as expressed).
#
# Our new limma pipeline reports fold-change on a log2 scale:
# positive values mean upregulated, negative values mean downregulated
# (e.g. log2(2) = 1, log2(0.5) = -1).
#
# To compare the two results on the same axis we log2-transform the old
# linear ratio. This puts both measures on an equivalent scale where
# 0 = no change, positive = up, negative = down, and the magnitude is
# directly comparable. We can verify the transformation is sensible:
# the old ratio range was 0.07 to 9.55, which becomes log2 ~-3.9 to 3.25,
# consistent with our new logFC range of -3.58 to 3.18.
results_joined <- results_joined |>
  dplyr::mutate(logFC_old = log2(ratio_old))

cat("Old ratio -> log2FC summary:\n")
print(summary(results_joined$logFC_old))
cat("\nNew logFC summary:\n")
print(summary(results_joined$logFC_new))
```

# Scatter plot: logFC old vs new

```{r fig.width = 7, fig.height = 6}
# Correlation coefficient for annotation on plot
r_val <- round(cor(results_joined$logFC_old,
                   results_joined$logFC_new,
                   use = "complete.obs"), 3)
scatter_plot <- ggplot(results_joined,
                       aes(x = logFC_old, y = logFC_new)) +
  geom_point(alpha = 0.3, size = 0.8, colour = "#2c7bb6") +
  geom_abline(intercept = 0, slope = 1,
              linetype   = "dashed",
              colour     = "firebrick",
              linewidth  = 0.7) +
  annotate("text",
           x    = -Inf, y = Inf,
           hjust = -0.15, vjust = 1.5,
           label = paste0("r = ", r_val),
           size  = 4.5) +
  labs(
    title    = "logFC comparison: new (limma) vs old (TAC) ANOVA",
    subtitle = paste0("Inner join on probe ID | n = ",
                      format(nrow(results_joined), big.mark = ","),
                      " probes"),
    x        = "Old results: log2(Ratio) [TAC]",
    y        = "New results: logFC [limma]",
    caption  = "Dashed red line = perfect concordance (slope = 1, intercept = 0)"
  ) +
  theme_bw(base_size = 13) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey40"),
        plot.caption  = element_text(colour = "grey50", size = 9))

print(scatter_plot)
```

# Save scatter plot

```{r}
ggsave(
  filename = file.path(out_dir, "scatter_logFC_new_vs_old.png"),
  plot     = scatter_plot,
  width    = 7,
  height   = 6,
  dpi      = 300
)
cat("Scatter plot saved.\n")
```

# Session info for reproducibility

```{r}
writeLines(capture.output(sessionInfo()),
           file.path(out_dir, "sessionInfo_comparison.txt"))

cat("Comparison complete.\n")
```


