---
title: "04-pathway_analysis_gata2_vs_con"
author: "Melanie Smith"
date: "2026-04-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Clean-up environment

```{r}
rm(list = ls(all.names = TRUE))
gc()
options(
  max.print              = .Machine$integer.max,
  scipen                 = 999,
  stringsAsFactors       = FALSE,
  dplyr.summarise.inform = FALSE
)
set.seed(42)
```

# Set project directories and file paths

```{r}
project_name  <- "20260421_natashaHarvey_gata2_mouse"
base_repo_dir <- "/home/melanie-smith/workDir/natashaHarvey"

project_repo_dir <- file.path(base_repo_dir, project_name)

processed_data <- file.path(project_repo_dir, "data", "processed")
out_dir        <- file.path(project_repo_dir, "outDir")
downloads_dir  <- file.path(project_repo_dir, "downloads")

# Pathway-specific output subdirectory to keep results organised
pathway_dir <- file.path(out_dir, "pathway_analysis")
dir.create(pathway_dir, recursive = TRUE, showWarnings = FALSE)

# Verify all required directories exist
for (d in c(processed_data, out_dir, downloads_dir)) {
  if (!dir.exists(d)) {
    stop("Required directory does not exist: ", d,
         "\nCreate it or update the path before re-running.")
  }
}
```

# Load libraries

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

# CRAN packages
cran_pkgs <- c("dplyr", "ggplot2", "scales", "RColorBrewer")
for (pkg in cran_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) install.packages(pkg)
  library(pkg, character.only = TRUE)
}

# Bioconductor packages
bioc_pkgs <- c(
  "clusterProfiler",   # ORA and GSEA framework
  "enrichplot",        # enrichment plots including GSEA running score
  "org.Mm.eg.db",      # Mouse gene annotation (Ensembl -> Entrez -> SYMBOL)
  "msigdbr",           # MSigDB gene sets (Hallmark, TFT)
  "DOSE"               # required by enrichplot
)
for (pkg in bioc_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    BiocManager::install(pkg, ask = FALSE, update = FALSE)
  }
  library(pkg, character.only = TRUE)
}
```

# Load DE results

```{r}
# Load full annotated DE results from script 03
results_annotated <- readRDS(file.path(processed_data,
                                       "lmFit_GATA2_vs_con_full.rds"))

cat("Loaded DE results:", nrow(results_annotated), "probes\n")
cat("Columns:", paste(colnames(results_annotated), collapse = ", "), "\n")

# Confirm required columns are present
required_cols <- c("PROBEID", "SYMBOL", "ENSEMBL", "logFC", "adj.P.Val", "t")
missing <- setdiff(required_cols, colnames(results_annotated))
if (length(missing) > 0) {
  stop("Missing required columns: ", paste(missing, collapse = ", "))
}
```

# Load independent expression filter (ania_eset_filt4.csv)

```{r}
# This file contains Transcript Cluster IDs from an independent microarray
# analysis of endothelial cells during embryonic development.
# Only probes present in this list are considered expressed in the relevant
# cell type and are retained for the ORA analysis.
#
# NOTE: To run pathway analysis WITHOUT this filter (second round),
# comment out the filter step in the ORA section below and set
# use_ania_filter <- FALSE
use_ania_filter <- TRUE

ania_filt <- read.csv(file.path(downloads_dir, "ania_eset_filt4.csv"),
                      stringsAsFactors = FALSE)

# Standardise column name — handle possible whitespace
colnames(ania_filt) <- trimws(colnames(ania_filt))

if (!"Transcript.Cluster.ID" %in% colnames(ania_filt)) {
  stop("Expected column 'Transcript Cluster ID' not found in ania_eset_filt4.csv.\n",
       "Available columns: ", paste(colnames(ania_filt), collapse = ", "))
}

ania_probes <- as.character(ania_filt[["Transcript.Cluster.ID"]])

cat("Probes in independent expression filter:", length(ania_probes), "\n")
cat("Probes in DE results matching filter:   ",
    sum(results_annotated$PROBEID %in% ania_probes), "\n")
```

# Compute rank metric for GSEA

```{r}
# Rank metric: sign(logFC) * -log10(adj.P.Val)
# Combines fold-change direction with statistical confidence.
# Strongly significant upregulated genes → large positive rank
# Strongly significant downregulated genes → large negative rank
# Genes with adj.P.Val close to 1 → rank near 0
#
# Guard against adj.P.Val == 0 which produces Inf
n_zero_pval <- sum(results_annotated$adj.P.Val == 0, na.rm = TRUE)
if (n_zero_pval > 0) {
  cat("WARNING:", n_zero_pval,
      "probes with adj.P.Val == 0 detected. Replacing with minimum non-zero value.\n")
  min_pval <- min(results_annotated$adj.P.Val[results_annotated$adj.P.Val > 0],
                  na.rm = TRUE)
  results_annotated$adj.P.Val[results_annotated$adj.P.Val == 0] <- min_pval
}

results_annotated$rank <- sign(results_annotated$logFC) *
                           (-log10(results_annotated$adj.P.Val))

cat("Rank metric summary:\n")
print(summary(results_annotated$rank))
```

# Resolve to one value per gene for pathway input

```{r}
# Pathway analysis requires one value per gene.
# For each SYMBOL, keep the probe with the most significant adj.P.Val.
# This is consistent with retaining the most statistically confident
# representative probe per gene.

# Remove probes with missing SYMBOL
results_clean <- results_annotated[!is.na(results_annotated$SYMBOL), ]

# Keep best probe per gene (lowest adj.P.Val)
results_gene <- results_clean[
  order(results_clean$SYMBOL, results_clean$adj.P.Val), ]
results_gene <- results_gene[!duplicated(results_gene$SYMBOL), ]

cat("Probes before gene-level collapse:", nrow(results_clean), "\n")
cat("Unique genes after collapse:      ", nrow(results_gene), "\n")
```

# Map gene identifiers to Entrez IDs

```{r}
# clusterProfiler and ReactomePA require Entrez IDs internally.
# Map from SYMBOL using org.Mm.eg.db (mouse annotation package).
symbol_to_entrez <- AnnotationDbi::select(
  org.Mm.eg.db,
  keys    = results_gene$SYMBOL,
  columns = c("ENTREZID", "SYMBOL"),
  keytype = "SYMBOL"
)

# Keep only unambiguous mappings
symbol_to_entrez <- symbol_to_entrez[!is.na(symbol_to_entrez$ENTREZID), ]
symbol_to_entrez <- symbol_to_entrez[!duplicated(symbol_to_entrez$SYMBOL), ]

# Merge Entrez IDs back into gene-level results
results_gene <- merge(results_gene, symbol_to_entrez, by = "SYMBOL", all.x = FALSE)

cat("Genes with Entrez ID mapping:", nrow(results_gene), "\n")
```

# ─────────────────────────────────────────────────────────────────────────────
# SECTION 1: OVER-REPRESENTATION ANALYSIS (ORA)
# Mirrors the original PANTHER analysis threshold:
#   FDR-adjusted p-value < 0.05 AND |logFC| > 1
# ─────────────────────────────────────────────────────────────────────────────

# Define ORA gene lists

```{r}
fdr_thresh <- 0.05
lfc_thresh <- 1

# Apply significance thresholds
sig_genes <- results_gene[
  results_gene$adj.P.Val < fdr_thresh &
    abs(results_gene$logFC) > lfc_thresh, ]

# ── ANIA FILTER (set use_ania_filter <- FALSE above to skip) ─────────────────
# Retain only probes expressed in embryonic endothelial cells (independent data)
# This mirrors the original PANTHER analysis exclusion criterion.
# To run without this filter, set use_ania_filter <- FALSE at the top.
if (use_ania_filter) {
  n_before <- nrow(sig_genes)
  sig_genes <- sig_genes[sig_genes$PROBEID %in% ania_probes, ]
  cat("Ania filter applied — probes before:", n_before,
      "| after:", nrow(sig_genes), "\n")
} else {
  cat("Ania filter NOT applied.\n")
}

# Separate up and down for directional ORA
sig_up   <- sig_genes[sig_genes$logFC > 0, ]
sig_down <- sig_genes[sig_genes$logFC < 0, ]

# Entrez ID vectors for ORA input
entrez_sig      <- sig_genes$ENTREZID
entrez_up       <- sig_up$ENTREZID
entrez_down     <- sig_down$ENTREZID
entrez_universe <- results_gene$ENTREZID   # all tested genes = background

cat("Significant genes (ORA input):", length(entrez_sig),
    "(", length(entrez_up), "up,", length(entrez_down), "down )\n")
cat("Universe (background):        ", length(entrez_universe), "genes\n")
```

# ORA helper function — shared plot style

```{r}
# Shared dotplot function for consistent ORA visualisation across databases
plot_ora_dotplot <- function(enrichment_result, title, n_show = 20) {
  if (is.null(enrichment_result) || nrow(enrichment_result) == 0) {
    message("No significant terms to plot for: ", title)
    return(invisible(NULL))
  }
  enrichplot::dotplot(
    enrichment_result,
    showCategory = n_show,
    font.size    = 9,
    title        = title
  ) +
    theme(plot.title = element_text(face = "bold", size = 12))
}

# Helper to save both the result CSV and the plot PNG
save_ora_results <- function(result_obj, prefix, plot_obj = NULL) {
  # Save CSV
  csv_path <- file.path(pathway_dir, paste0(prefix, "_results.csv"))
  write.csv(as.data.frame(result_obj), file = csv_path, row.names = FALSE)
  cat("Results saved:", csv_path, "\n")

  # Save plot
  if (!is.null(plot_obj)) {
    png_path <- file.path(pathway_dir, paste0(prefix, "_dotplot.png"))
    png(png_path, width = 10, height = 8, units = "in", res = 300)
    print(plot_obj)
    dev.off()
    cat("Plot saved:   ", png_path, "\n")
  }
}
```

# ORA 1: GO Biological Process

```{r ora_go_bp, fig.width = 10, fig.height = 8}
ora_go_bp <- clusterProfiler::enrichGO(
  gene          = entrez_sig,
  universe      = entrez_universe,
  OrgDb         = org.Mm.eg.db,
  ont           = "BP",             # Biological Process
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2,
  readable      = TRUE              # convert Entrez IDs back to gene symbols
)

# Simplify to remove highly redundant GO terms
ora_go_bp_simplified <- clusterProfiler::simplify(
  ora_go_bp,
  cutoff    = 0.7,
  by        = "p.adjust",
  select_fun = min
)

cat("GO BP significant terms (before simplify):", nrow(ora_go_bp),      "\n")
cat("GO BP significant terms (after simplify): ", nrow(ora_go_bp_simplified), "\n")

p_ora_go_bp <- plot_ora_dotplot(
  ora_go_bp_simplified,
  title  = paste0("ORA — GO Biological Process\nGATA2 vs control",
                  ifelse(use_ania_filter, " (ania filtered)", ""))
)
print(p_ora_go_bp)

save_ora_results(ora_go_bp_simplified, "ORA_GO_BP", p_ora_go_bp)
```

# ORA 2: KEGG

```{r ora_kegg, fig.width = 10, fig.height = 8}
ora_kegg <- clusterProfiler::enrichKEGG(
  gene          = entrez_sig,
  universe      = entrez_universe,
  organism      = "mmu",            # Mus musculus
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2
)

# Convert Entrez IDs to readable gene symbols in output
ora_kegg <- clusterProfiler::setReadable(ora_kegg,
                                         OrgDb   = org.Mm.eg.db,
                                         keyType = "ENTREZID")

cat("KEGG significant pathways:", nrow(ora_kegg), "\n")

p_ora_kegg <- plot_ora_dotplot(
  ora_kegg,
  title = paste0("ORA — KEGG Pathways\nGATA2 vs control",
                 ifelse(use_ania_filter, " (ania filtered)", ""))
)
print(p_ora_kegg)

save_ora_results(ora_kegg, "ORA_KEGG", p_ora_kegg)
```

# ORA 3: Reactome

```{r ora_reactome, fig.width = 10, fig.height = 8}
# ORA 3: Reactome (via MSigDB C2 Reactome gene sets)
msig_reactome <- msigdbr::msigdbr(species = "Mus musculus",
                                   category = "C2",
                                   subcategory = "CP:REACTOME") |>
  dplyr::select(gs_name, entrez_gene) |>
  dplyr::mutate(entrez_gene = as.character(entrez_gene))

ora_reactome <- clusterProfiler::enricher(
  gene          = entrez_sig,
  universe      = entrez_universe,
  TERM2GENE     = msig_reactome,
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2
)

ora_reactome <- clusterProfiler::setReadable(ora_reactome,
                                              OrgDb   = org.Mm.eg.db,
                                              keyType = "ENTREZID")

cat("Reactome (MSigDB C2) significant pathways:", nrow(ora_reactome), "\n")

p_ora_reactome <- plot_ora_dotplot(
  ora_reactome,
  title = paste0("ORA — Reactome Pathways (MSigDB C2)\nGATA2 vs control",
                 ifelse(use_ania_filter, " (ania filtered)", ""))
)
print(p_ora_reactome)

save_ora_results(ora_reactome, "ORA_Reactome", p_ora_reactome)
```

# ORA 4: MSigDB Hallmark

```{r ora_hallmark, fig.width = 10, fig.height = 8}
# Retrieve mouse Hallmark gene sets from MSigDB
msig_hallmark <- msigdbr::msigdbr(species = "Mus musculus",
                                   category = "H") |>
  dplyr::select(gs_name, entrez_gene) |>
  dplyr::mutate(entrez_gene = as.character(entrez_gene))

ora_hallmark <- clusterProfiler::enricher(
  gene          = entrez_sig,
  universe      = entrez_universe,
  TERM2GENE     = msig_hallmark,
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2
)

ora_hallmark <- clusterProfiler::setReadable(ora_hallmark,
                                              OrgDb   = org.Mm.eg.db,
                                              keyType = "ENTREZID")

cat("MSigDB Hallmark significant gene sets:", nrow(ora_hallmark), "\n")

p_ora_hallmark <- plot_ora_dotplot(
  ora_hallmark,
  title = paste0("ORA — MSigDB Hallmark Gene Sets\nGATA2 vs control",
                 ifelse(use_ania_filter, " (ania filtered)", ""))
)
print(p_ora_hallmark)

save_ora_results(ora_hallmark, "ORA_MSigDB_Hallmark", p_ora_hallmark)
```

# ORA 5: MSigDB C3 Transcription Factor Targets (TFT)

```{r ora_tft, fig.width = 10, fig.height = 8}
# C3 TFT — transcription factor target gene sets
# Particularly relevant for a GATA2 knockdown experiment
msig_tft <- msigdbr::msigdbr(species = "Mus musculus",
                               category = "C3",
                               subcategory = "TFT:GTRD") |>
  dplyr::select(gs_name, entrez_gene) |>
  dplyr::mutate(entrez_gene = as.character(entrez_gene))

ora_tft <- clusterProfiler::enricher(
  gene          = entrez_sig,
  universe      = entrez_universe,
  TERM2GENE     = msig_tft,
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2
)

ora_tft <- clusterProfiler::setReadable(ora_tft,
                                         OrgDb   = org.Mm.eg.db,
                                         keyType = "ENTREZID")

cat("MSigDB C3 TFT significant gene sets:", nrow(ora_tft), "\n")

p_ora_tft <- plot_ora_dotplot(
  ora_tft,
  title = paste0("ORA — MSigDB C3 Transcription Factor Targets\nGATA2 vs control",
                 ifelse(use_ania_filter, " (ania filtered)", ""))
)
print(p_ora_tft)

save_ora_results(ora_tft, "ORA_MSigDB_TFT", p_ora_tft)
```

# ─────────────────────────────────────────────────────────────────────────────
# SECTION 2: GENE SET ENRICHMENT ANALYSIS (GSEA)
# Uses ALL genes ranked by: sign(logFC) * -log10(adj.P.Val)
# No significance threshold applied — the full ranked list is used
# ─────────────────────────────────────────────────────────────────────────────

# Build ranked gene vector for GSEA

```{r}
# Named numeric vector: names = Entrez IDs, values = rank metric
# Must be sorted in decreasing order for clusterProfiler::GSEA()
ranked_vec <- results_gene$rank
names(ranked_vec) <- results_gene$ENTREZID

# Remove any NA names or values
ranked_vec <- ranked_vec[!is.na(names(ranked_vec)) & !is.na(ranked_vec)]

# Sort descending
ranked_vec <- sort(ranked_vec, decreasing = TRUE)

cat("Ranked gene vector length:", length(ranked_vec), "\n")
cat("Rank metric range: [", round(min(ranked_vec), 3),
    ",", round(max(ranked_vec), 3), "]\n")
```

# GSEA helper functions

```{r}
# Shared GSEA dotplot
plot_gsea_dotplot <- function(gsea_result, title, n_show = 20) {
  if (is.null(gsea_result) || nrow(gsea_result) == 0) {
    message("No significant terms to plot for: ", title)
    return(invisible(NULL))
  }
  enrichplot::dotplot(
    gsea_result,
    showCategory = n_show,
    split        = ".sign",       # split by enriched/suppressed
    font.size    = 9,
    title        = title
  ) +
    facet_grid(. ~ .sign) +
    theme(plot.title = element_text(face = "bold", size = 12))
}

# GSEA enrichment plot for top N terms
plot_gsea_enrichment <- function(gsea_result, prefix, n_top = 5) {
  if (is.null(gsea_result) || nrow(gsea_result) == 0) {
    message("No terms to plot enrichment for: ", prefix)
    return(invisible(NULL))
  }

  result_df <- as.data.frame(gsea_result)

  # Top N by NES (most positive and most negative)
  top_pos <- result_df |>
    dplyr::filter(NES > 0) |>
    dplyr::slice_max(order_by = NES, n = n_top)
  top_neg <- result_df |>
    dplyr::filter(NES < 0) |>
    dplyr::slice_min(order_by = NES, n = n_top)
  top_ids <- c(top_pos$ID, top_neg$ID)

  # Plot and save each enrichment plot
  for (id in top_ids) {
    term_name <- result_df$Description[result_df$ID == id]
    safe_name <- gsub("[^A-Za-z0-9_]", "_", term_name)
    p <- enrichplot::gseaplot2(
      gsea_result,
      geneSetID = id,
      title     = term_name,
      pvalue_by = "NES"
    )
    png_path <- file.path(pathway_dir,
                          paste0("GSEA_", prefix, "_enrichplot_",
                                 safe_name, ".png"))
    png(png_path, width = 10, height = 6, units = "in", res = 300)
    print(p)
    dev.off()
    cat("Enrichment plot saved:", png_path, "\n")
  }
}

# Save GSEA CSV and dotplot
save_gsea_results <- function(result_obj, prefix, plot_obj = NULL) {
  csv_path <- file.path(pathway_dir, paste0(prefix, "_results.csv"))
  write.csv(as.data.frame(result_obj), file = csv_path, row.names = FALSE)
  cat("Results saved:", csv_path, "\n")

  if (!is.null(plot_obj)) {
    png_path <- file.path(pathway_dir, paste0(prefix, "_dotplot.png"))
    png(png_path, width = 10, height = 8, units = "in", res = 300)
    print(plot_obj)
    dev.off()
    cat("Plot saved:   ", png_path, "\n")
  }
}
```

# GSEA 1: GO Biological Process

```{r gsea_go_bp, fig.width = 10, fig.height = 8}
gsea_go_bp <- clusterProfiler::gseGO(
  geneList      = ranked_vec,
  OrgDb         = org.Mm.eg.db,
  ont           = "BP",
  keyType       = "ENTREZID",
  minGSSize     = 15,
  maxGSSize     = 500,
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  verbose       = FALSE,
  seed          = TRUE      # reproducibility
)

gsea_go_bp <- clusterProfiler::setReadable(gsea_go_bp,
                                            OrgDb   = org.Mm.eg.db,
                                            keyType = "ENTREZID")

cat("GSEA GO BP significant terms:", nrow(gsea_go_bp), "\n")

p_gsea_go_bp <- plot_gsea_dotplot(gsea_go_bp,
                                   title = "GSEA — GO Biological Process\nGATA2 vs control")
print(p_gsea_go_bp)

save_gsea_results(gsea_go_bp, "GSEA_GO_BP", p_gsea_go_bp)
plot_gsea_enrichment(gsea_go_bp, prefix = "GO_BP", n_top = 5)
```

# GSEA 2: KEGG

```{r gsea_kegg, fig.width = 10, fig.height = 8}
gsea_kegg <- clusterProfiler::gseKEGG(
  geneList      = ranked_vec,
  organism      = "mmu",
  minGSSize     = 15,
  maxGSSize     = 500,
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  verbose       = FALSE,
  seed          = TRUE
)

gsea_kegg <- clusterProfiler::setReadable(gsea_kegg,
                                           OrgDb   = org.Mm.eg.db,
                                           keyType = "ENTREZID")

cat("GSEA KEGG significant pathways:", nrow(gsea_kegg), "\n")

p_gsea_kegg <- plot_gsea_dotplot(gsea_kegg,
                                  title = "GSEA — KEGG Pathways\nGATA2 vs control")
print(p_gsea_kegg)

save_gsea_results(gsea_kegg, "GSEA_KEGG", p_gsea_kegg)
plot_gsea_enrichment(gsea_kegg, prefix = "KEGG", n_top = 5)
```

# GSEA 3: Reactome

```{r gsea_reactome, fig.width = 10, fig.height = 8}
# GSEA 3: Reactome (via MSigDB C2 Reactome gene sets)
gsea_reactome <- clusterProfiler::GSEA(
  geneList      = ranked_vec,
  TERM2GENE     = msig_reactome,   # reuses object created in ORA chunk above
  minGSSize     = 15,
  maxGSSize     = 500,
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  verbose       = FALSE,
  seed          = TRUE
)

gsea_reactome <- clusterProfiler::setReadable(gsea_reactome,
                                               OrgDb   = org.Mm.eg.db,
                                               keyType = "ENTREZID")

cat("GSEA Reactome (MSigDB C2) significant pathways:", nrow(gsea_reactome), "\n")

p_gsea_reactome <- plot_gsea_dotplot(gsea_reactome,
                                      title = "GSEA — Reactome Pathways (MSigDB C2)\nGATA2 vs control")
print(p_gsea_reactome)

save_gsea_results(gsea_reactome, "GSEA_Reactome", p_gsea_reactome)
plot_gsea_enrichment(gsea_reactome, prefix = "Reactome", n_top = 5)
```

# GSEA 4: MSigDB Hallmark

```{r gsea_hallmark, fig.width = 10, fig.height = 8}
msig_hallmark_gsea <- msigdbr::msigdbr(species = "Mus musculus",
                                        category = "H") |>
  dplyr::select(gs_name, entrez_gene) |>
  dplyr::mutate(entrez_gene = as.character(entrez_gene))

gsea_hallmark <- clusterProfiler::GSEA(
  geneList      = ranked_vec,
  TERM2GENE     = msig_hallmark_gsea,
  minGSSize     = 15,
  maxGSSize     = 500,
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  verbose       = FALSE,
  seed          = TRUE
)

gsea_hallmark <- clusterProfiler::setReadable(gsea_hallmark,
                                               OrgDb   = org.Mm.eg.db,
                                               keyType = "ENTREZID")

cat("GSEA MSigDB Hallmark significant gene sets:", nrow(gsea_hallmark), "\n")

p_gsea_hallmark <- plot_gsea_dotplot(gsea_hallmark,
                                      title = "GSEA — MSigDB Hallmark Gene Sets\nGATA2 vs control")
print(p_gsea_hallmark)

save_gsea_results(gsea_hallmark, "GSEA_MSigDB_Hallmark", p_gsea_hallmark)
plot_gsea_enrichment(gsea_hallmark, prefix = "Hallmark", n_top = 5)
```

# GSEA 5: MSigDB C3 Transcription Factor Targets

```{r gsea_tft, fig.width = 10, fig.height = 8}
msig_tft_gsea <- msigdbr::msigdbr(species = "Mus musculus",
                                   category = "C3",
                                   subcategory = "TFT:GTRD") |>
  dplyr::select(gs_name, entrez_gene) |>
  dplyr::mutate(entrez_gene = as.character(entrez_gene))

gsea_tft <- clusterProfiler::GSEA(
  geneList      = ranked_vec,
  TERM2GENE     = msig_tft_gsea,
  minGSSize     = 15,
  maxGSSize     = 500,
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  verbose       = FALSE,
  seed          = TRUE
)

gsea_tft <- clusterProfiler::setReadable(gsea_tft,
                                          OrgDb   = org.Mm.eg.db,
                                          keyType = "ENTREZID")

cat("GSEA MSigDB C3 TFT significant gene sets:", nrow(gsea_tft), "\n")

p_gsea_tft <- plot_gsea_dotplot(gsea_tft,
                                 title = "GSEA — MSigDB C3 Transcription Factor Targets\nGATA2 vs control")
print(p_gsea_tft)

save_gsea_results(gsea_tft, "GSEA_MSigDB_TFT", p_gsea_tft)
plot_gsea_enrichment(gsea_tft, prefix = "TFT", n_top = 5)
```

# Summary of all pathway results

```{r}
# Compile a brief summary of significant terms found across all analyses
summary_tbl <- data.frame(
  Analysis  = c("ORA — GO BP", "ORA — KEGG", "ORA — Reactome",
                 "ORA — Hallmark", "ORA — TFT",
                 "GSEA — GO BP", "GSEA — KEGG", "GSEA — Reactome",
                 "GSEA — Hallmark", "GSEA — TFT"),
  Method    = c(rep("ORA", 5), rep("GSEA", 5)),
  Database  = c("GO BP", "KEGG", "Reactome", "MSigDB Hallmark", "MSigDB C3 TFT",
                 "GO BP", "KEGG", "Reactome", "MSigDB Hallmark", "MSigDB C3 TFT"),
  N_sig     = c(
    nrow(ora_go_bp_simplified), nrow(ora_kegg),
    nrow(ora_reactome),         nrow(ora_hallmark),
    nrow(ora_tft),
    nrow(gsea_go_bp),           nrow(gsea_kegg),
    nrow(gsea_reactome),        nrow(gsea_hallmark),
    nrow(gsea_tft)
  )
)

cat("=== Pathway analysis summary ===\n")
print(summary_tbl)

write.csv(summary_tbl,
          file      = file.path(pathway_dir, "pathway_analysis_summary.csv"),
          row.names = FALSE)

cat("\nSummary table saved to:",
    file.path(pathway_dir, "pathway_analysis_summary.csv"), "\n")
```

# Session info for reproducibility

```{r}
writeLines(
  capture.output(sessionInfo()),
  file.path(out_dir, "sessionInfo_pathway_analysis.txt")
)

cat("Pathway analysis complete.\n")
cat("All results saved to:", pathway_dir, "\n")
```
